\documentclass[fsharpNotes.tex]{subfiles}
\graphicspath{ {./figures/} }

\begin{document}
\chapter{To Do}
\begin{itemize}
\item Make sure \lstinline{#quit;;} or \texttt{ctrl-d} is included in the first chapters
\end{itemize}
\chapter{Old To Dos}
\begin{itemize}
\item Go through all idxss and make sure they should not be in the
  margin.
\item Figure out a consistent way to handle System... in the margin
  and index.
\item Integrate language details into main text.
\item Add Winforms Bitmap (and fillable objects)
\item Add Winforms Refresh
\item Add appendix on regular expressions
\item Add Torben's notes on functional programming
\item Add a chapter comparing the 3 paradigms
\item Add abstraction of computer: places <-> memory/disk. Mutable objects are abstractions of places \url{https://www.infoq.com/presentations/Value-Values}. Facts does not rime with set and get. 
\item Hickey: Difference between syntax and semantics. Values or locations, add a good figure. Functional programming: All values are freely shareable.
\item something about organizing stuff: \url{https://fsharpforfunandprofit.com/posts/organizing-functions/}, \url{https://fsharpforfunandprofit.com/posts/recipe-part3/}
\item why discriminant unions and function overloading is problematic: \url{http://stefanalfbo.github.io/blog/2014/01/05/fsharp-function-overloading/}
\item Maybe something from \url{https://fsharpforfunandprofit.com/posts/type-extensions/}
\item Maybe publish with \url{https://www.lulu.com}
\item consider following convention \url{http://fsprojects.github.io/FSharpLint/NameConventions.html},  particularly PascalCase instead of camelCase.
\item Somewhere describe generic functions and type definitions
\item Credit to Hans for early version of Console chapter.
\item Possible highlight changes in code stumps, and possibly improve reference to code used for each compilation example.
\item Railway oriented programming \url{https://fsharpforfunandprofit.com/rop/}
\item Reference and value discussion in listable things.
\item Possibly an overview table of reference and value types.
\item Some possible oop examples \url{https://www3.ntu.edu.sg/home/ehchua/programming/cpp/cp5_OOPExamples.html}
\item Should boxing be mentioned? \lstinline|(box 5) :?> int;;|, see Spec-4.0 chapter 18.2.6.
\item In object-oriented programming: functions and data are combined. Contrast the Anemic Domain Model (\url{https://www.martinfowler.com/bliki/AnemicDomainModel.html})
\item boxing \lstinline|(box 5) :?> int;;|, see Spec-4.0 chapter 18.2.6.
\end{itemize}
\end{document}
%%% Local Variables:
%%% TeX-master: "fsharpNotes"
%%% End:

